Finch: Sparse and Structured Array Programming with Control Flow

### 2024-03-31 sections/background.tex
Update on Overleaf.

In this work, we partially simplify the presentation of Looplets to focus on the
semantics, rather than precise implementation. Since Finch is designed to be
embedded into another language, we allow for several of the looplets to utilize
functions that modify the state of that language. It is assumed that if a
looplet introduces a variable to be used in a child looplet, the child looplet
will not modify that variaable.

>>>>

In this work, we partially simplify the presentation of Looplets to focus on the
semantics, rather than precise implementation. Since Finch is designed to be
embedded into another language\saman{not clear}, we allow for several of the looplets to utilize
functions that modify the state of that language. It is assumed that if a
looplet introduces a variable to be used in a child looplet, the child looplet
will not modify that variable.\saman{No context yet to understand this. Need to introduce what looplets are (Fig 2)!}

### 2024-03-31 sections/eval.tex
Update on Overleaf.
yay

% TODO: make this be 3 columns
\begin{figure}
    \begin{minipage}[t]{0.315\textwidth}
        \vspace{0pt} % Add this to ensure top alignment within minipage
        \begin{minted}{julia}
            @finch begin
                y .= 0
                for j = _, i = _
                    y[i] += A[i, j] * x[j]
                end
                return y
            end
        \end{minted}
    \end{minipage}%
    \begin{minipage}[t]{0.315\textwidth}
        \vspace{0pt} % Add this to ensure top alignment within minipage
        \begin{minted}{julia}
            @finch begin
                y .= 0
                for j = _, i = _
                    y[j] += A[i, j] * x[i]
                end
                return y
            end
        \end{minted}
    \end{minipage}
    \begin{minipage}[t]{0.36\textwidth}
        \vspace{0pt} % Add this to ensure top alignment within minipage
        \begin{minted}{julia}
            @finch begin
                y .= 0
                for j = _
                    let x_j = x[j]
                        y_j .= 0
                        for i = _
                            let A_ij = A[i, j]
                                y[i] += x_j * A_ij
                                y_j[] += A_ij * x[i]
                            end
                        end
                        y[j] += y_j[] + diag[j] * x_j
                    end
                end
                return y
            end
        \end{minted}
    \end{minipage}
    \caption{Finch SpMV Programs}
    \label{spmv_programs}
\end{figure}

\begin{table}[htbp]
    \centering
    \caption{SpMV Tensor Formats}
    \label{spmv_tensor_formats}
    \begin{tabular}{|l|l|l|l|}
        \hline
        \textbf{Outer Level} & \textbf{Inner Level} & \textbf{Scalar Values} & \textbf{Style of Matrix}\\
        \hline
        \multirow{6}{*}{Dense} & \multirow{2}{*}{SparseList} & Element & sparse, real-valued matrices \\
        \cline{3-4} 
        & & Pattern & sparse, boolean-valued matrices \\
        \cline{2-4} 
        & SparseVBL & Element & real-valued matrices with blocked structure \\
        \cline{2-4}
        & SparseBand & Element & real-valued matrices with diagonal band \\
        \cline{2-4}
        & \multirow{2}{*}{SparsePoint} & Element & real-valued matrices with one value per row \\
        \cline{3-4} 
        & & Pattern & matrix with runs of true or false \\
        \hline 
    \end{tabular}
\end{table}

@@

\begin{table}[htbp]
    \centering
    \caption{Transposed SpMV Sample Matrices}
    \label{tab:transposed_spmv_sample_matrices}
    \begin{tabular}{|
        >{\centering\arraybackslash}m{0.2\linewidth-2\tabcolsep-1.2\arrayrulewidth}|
        >{\centering\arraybackslash}m{0.2\linewidth-2\tabcolsep-1.2\arrayrulewidth}|
        >{\centering\arraybackslash}m{0.2\linewidth-2\tabcolsep-1.2\arrayrulewidth}|
        >{\centering\arraybackslash}m{0.2\linewidth-2\tabcolsep-1.2\arrayrulewidth}|
        >{\centering\arraybackslash}m{0.2\linewidth-2\tabcolsep-1.2\arrayrulewidth}|}
        \hline
         & \textbf{HB/saylr4} & \textbf{Norris/heart3} & \textbf{large\_band} & \textbf{SNAP/as-735} \\
        \hline
        \textbf{Spy} & \includegraphics[width=\linewidth]{spmv_matrices/saylr4.png} & \includegraphics[width=\linewidth]{spmv_matrices/heart3.png} & \includegraphics[width=\linewidth]{spmv_matrices/toeplitz_large_band.png} & \includegraphics[width=\linewidth]{spmv_matrices/as-735.png} \\
        \hline 
        \textbf{Group / Name} & HB/saylr4 & Norris/heart3 & large\_band & SNAP/as-735 \\
        \hline
        \textbf{Dimensions} & 3,564 x 3,564 (22,316) & 2,339 x 2,339 (680,341) & 10,000 x 10,000 (1,999,900) & 7,716 x 7,716 (26,467) \\
        \hline
        \textbf{Best Finch Format} & Symmetric SparseList & SparseVBL & SparseBand & Symmetric SparseList-Pattern \\
        \hline
    \end{tabular}
\end{table}

@@

\begin{figure}
    \includegraphics[width=\linewidth]{Spmv_annotated.drawio.png}
    \caption{Performance of SpMV by Finch format.}
    \label{fig:spmv_grouped}
    \footnotesize The performance displayed for Finch on each dataset is the fastest among the formats we tested. "R" indicates row-major implementation and "C" indicates column-major implementation in Finch. The baseline Finch format is unsymmetric Dense(SparseList(Element)).
\end{figure}

@@

\subsubsection{Finch Interpreter} 
Once the Finch Logic program has been converted to the standard form, the Finch Interpreter can execute each query, in order, through a straightforward lowering process. First, the output format is identified by unpacking the outer $\finchreformat$ statement. Next, the inner expression of the is unpacked to identify the aggregate operator and to convert the pointwise expression into a Finch expression. At this point, any aliases to the result of previous queries are replaced with an access to the actual tensor. Lastly, the concordant loop order is identified and instantiated. The lowered query can then be compiled and executed with the Finch compiler, and the result is assigned to $name$ in the plan's scope before proceeding to the next query in the plan.

>>>>

% TODO: make this be 3 columns
\begin{figure}
    \begin{minipage}[t]{0.315\textwidth}
        \vspace{0pt} % Add this to ensure top alignment within minipage
        \begin{minted}{julia}
            @finch begin
                y .= 0
                for j = _, i = _
                    y[i] += A[i, j] * x[j]
                end
                return y
            end
        \end{minted}
    \end{minipage}%
    \begin{minipage}[t]{0.315\textwidth}
        \vspace{0pt} % Add this to ensure top alignment within minipage
        \begin{minted}{julia}
            @finch begin
                y .= 0
                for j = _, i = _
                    y[j] += A[i, j] * x[i]
                end
                return y
            end
        \end{minted}
    \end{minipage}
    \begin{minipage}[t]{0.36\textwidth}
        \vspace{0pt} % Add this to ensure top alignment within minipage
        \begin{minted}{julia}
            @finch begin
                y .= 0
                for j = _
                    let x_j = x[j]
                        y_j .= 0
                        for i = _
                            let A_ij = A[i, j]
                                y[i] += x_j * A_ij
                                y_j[] += A_ij * x[i]
                            end
                        end
                        y[j] += y_j[] + diag[j] * x_j
                    end
                end
                return y
            end
        \end{minted}
    \end{minipage}
    \caption{Finch \saman{row-major, column-major and symmetric} SpMV Programs}
    \label{spmv_programs}
\end{figure}

\begin{table}[htbp]
    \footnotesize
    \centering
    \caption{SpMV Tensor Formats}
    \label{spmv_tensor_formats}
    \begin{tabular}{|l|l|l|l|}
        \hline
        \textbf{Outer Level} & \textbf{Inner Level} & \textbf{Scalar Values} & \textbf{Style of Matrix}\\
        \hline
        \multirow{6}{*}{Dense} & \multirow{2}{*}{SparseList} & Element & sparse, real-valued matrices \\
        \cline{3-4} 
        & & Pattern & sparse, boolean-valued matrices \\
        \cline{2-4} 
        & SparseVBL & Element & real-valued matrices with blocked structure \\
        \cline{2-4}
        & SparseBand & Element & real-valued matrices with diagonal band \\
        \cline{2-4}
        & \multirow{2}{*}{SparsePoint} & Element & real-valued matrices with one value per row \\
        \cline{3-4} 
        & & Pattern & matrix with runs of true or false \\
        \hline 
    \end{tabular}
\end{table}

@@

\begin{table}[htbp]
    \centering
    \footnotesize
    \caption{Transposed SpMV Sample Matrices}
    \label{tab:transposed_spmv_sample_matrices}
    \begin{tabular}{|
        >{\centering\arraybackslash}m{0.2\linewidth-2\tabcolsep-1.2\arrayrulewidth}|
        >{\centering\arraybackslash}m{0.2\linewidth-2\tabcolsep-1.2\arrayrulewidth}|
        >{\centering\arraybackslash}m{0.2\linewidth-2\tabcolsep-1.2\arrayrulewidth}|
        >{\centering\arraybackslash}m{0.2\linewidth-2\tabcolsep-1.2\arrayrulewidth}|
        >{\centering\arraybackslash}m{0.2\linewidth-2\tabcolsep-1.2\arrayrulewidth}|}
        \hline
         & \textbf{HB/saylr4} & \textbf{Norris/heart3} & \textbf{large\_band} & \textbf{SNAP/as-735} \\
        \hline
        \textbf{Spy} & \includegraphics[width=\linewidth]{spmv_matrices/saylr4.png} & \includegraphics[width=\linewidth]{spmv_matrices/heart3.png} & \includegraphics[width=\linewidth]{spmv_matrices/toeplitz_large_band.png} & \includegraphics[width=\linewidth]{spmv_matrices/as-735.png} \\
        \hline 
        \textbf{Group / Name} & HB/saylr4 & Norris/heart3 & large\_band & SNAP/as-735 \\
        \hline
        \textbf{Dimensions} & 3,564 x 3,564 (22,316) & 2,339 x 2,339 (680,341) & 10,000 x 10,000 (1,999,900) & 7,716 x 7,716 (26,467) \\
        \hline
        \textbf{Best Finch Format} & Symmetric SparseList & SparseVBL & SparseBand & Symmetric SparseList-Pattern \\
        \hline
    \end{tabular}
\end{table}

@@

\begin{figure}
    \includegraphics[width=\linewidth]{Spmv_annotated.png}
    \caption{Performance of SpMV by Finch format.}
    \label{fig:spmv_grouped}
    \footnotesize The performance displayed for Finch on each dataset is the fastest among the formats we tested. "R" indicates row-major implementation and "C" indicates column-major implementation in Finch. The baseline Finch format is unsymmetric Dense(SparseList(Element)).
\end{figure}

@@

\subsubsection{Finch Interpreter} 
Once the Finch Logic program has been converted to the standard form, the Finch Interpreter can execute each query, in order, through a straightforward lowering process. First, the output format is identified by unpacking the outer $\finchreformat$ statement. Next, the inner expression of the is unpacked to identify the $\finchaggregate$ operator and to convert the pointwise expression into a Finch expression. At this point, any aliases to the result of previous queries are replaced with an access to the actual tensor. Lastly, the concordant loop order is identified and instantiated. The lowered query can then be compiled and executed with the Finch compiler, and the result is assigned to $name$ in the plan's scope before proceeding to the next query in the plan.

### 2024-03-31 sections/intro.tex
Update on Overleaf.

%Array programming is the core abstraction behind many of the modern miracles of computing (e.g. neural networks, scientific simulation, database processing). 
Arrays are the most fundamental abstraction in computer science. Arrays and lists are often the first-taught datastructure
\cite[Chapter 2.2]{abelson_structure_1996}, \cite[Chapter 2.2]{knuth_art_1997}.
%
Arrays are also universal across programming languages, from their introduction
in Fortran in 1957 to present-day languages like Python
\cite{backus_fortran_1957}, keeping more-or-less the same semantics.
%
Modern
array programming languages such as NumPy, SciPy, MatLab, TensorFlow, PyTorch,
and Halide have pushed the limits of productive data processing with arrays,
fueling breakthroughs in machine learning, scientific computing, image
processing, and more  \cite{harris_array_2020, virtanen_scipy_2020,
moler_history_2020, abadi_tensorflow_2016,
paszke_pytorch_2019,ragan-kelley_halide_2013}.
% EASY CUT
%These frameworks have been the
%subject of extensive industry investment to enable performant implementations,
%and often operate at the peak capacity of the hardware they run on
%\cite{lo_roofline_2015}.

The success and ubiquity of arrays is largely due to their simplicity. 
%
Since their introduction, multidimensional arrays have represented dense, rectilinear,
integer grids of points. 
%
By \textbf{dense}, we mean that indices are mapped to value via a simple formula relating multidimensional space to linear memory.
%
Consequently, dense arrays offer extensive compiler optimizations and many convenient interfaces.
%
%This
%simplicity enables extensive interoperability, convenience layers, and
%optimizations by breaking the abstraction barrier between array representation
%and array storage.  
%
Compilers understand dense array computations across many
programming constructs, such as for and while loops, breaks, parallelism,
caching, prefetching, multiple outputs, scatters, gathers, vectorization,
loop-carry-dependencies, and more. Several optimizations have been developed for
dense arrays, such as loop fusion, loop tiling, loop unrolling, and loop
interchange.
%
However, while dense arrays are the easiest way to program for performance, the world is not all dense.

@@

%
\textbf{Currently, support for structured data is fragmented and incomplete}.
%
Experts must hand write variations of even the simplest kernels, like matrix
multiply, for each data structure/data set and architecture to get performance.
%
Implementations must choose a small set of features to support well, resulting
in a compromise between \textbf{program flexibility} and \textbf{data structure
flexibility}.
%
Hand-written solutions are collected in diverse libraries like
MKL, OpenCV, LAPACK or SciPy~\cite{ bradski2000opencv, anderson1999lapack, virtanen2020scipy, psarras2022linear}. 
%
However, libraries will only ever support a subset of
programs on a subset of data structure combinations.
%
Even the most advanced
libraries, such as the GraphBLAS, which support a wide variety of sparse
operations over various semi-rings always lack support for other features, such
as tensors, fused outputs, or runs of repeated values~\cite{bulucc2017design, mattson2019lagraph}.
%
While dense array
compilers support an enormous variety of program constructs like early break and
multiple left hand sides, they only support dense arrays~\cite{ragan-kelley_halide_2013,grosser2012polly}.  
%
Special-purpose
compilers like TACO, Taichi, StructTensor, or CoRa which support a select subset of structured data
structures (only sparse, or only ragged arrays) must compromise by greatly
constraining the classes of programs which they support, such as tensor
contractions \cite{kjolstad_tensor_2019, hu_taichi_2019, ghorbani2023compiling, fegade_cora_2022}. 
%
This trade-off is visualized in Tables \ref{tab:features} and \ref{tab:data_structures}.
%

@@

\begin{table}[h!]
\noindent
\flushleft
\begin{minipage}[t]{.49\textwidth}
  \scriptsize
  \begin{tabular}{l|cccccc}
  \textbf{Feature / Tool} & \rothead{Halide} & \rothead{Taco} & \rothead{Cora} & \rothead{Taichi} & \rothead{Stur} & \rothead{Finch} \\
  \hline
  Einsums/Contractions & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
  % High-Level API           &            & \checkmark &            &            &            & \checkmark \\
  % Automatic API Fusion     &            &            &            &            &            & \checkmark \\
%  Parallelism             & \checkmark & \checkmark & \checkmark & \checkmark&           & \checkmark \\
  Multiple LHS             & \checkmark &            & \checkmark & \checkmark &            & \checkmark \\
  Affine Indices           & \checkmark &            &            & \checkmark & \checkmark & \checkmark \\
  Recurrence               & \checkmark &            &            &            &            &           \\
  If-Conditions and Masks  & \checkmark & \checkmark &            & \checkmark &            & \checkmark \\
  Scatter Gather           & \checkmark &            &            & \checkmark &            &\checkmark \\
  Early Break              &            & \checkmark &            & \checkmark &            &\checkmark \\
  Unrestricted Read/Write              &    \checkmark        &  &            &  &            &  \\
  \end{tabular}
  \caption{Feature support across various tools.}
  \label{tab:features}
\end{minipage} 
\begin{minipage}[t]{.49\textwidth}
  \flushright
  \scriptsize
  \begin{tabular}{l|cccccc}
  \textbf{Feature / Tool} & \rothead{Halide} & \rothead{Taco} & \rothead{Cora} & \rothead{Taichi} & \rothead{Stur} & \rothead{Finch} \\
  \hline
  Dense                    & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
  Padded                   & \checkmark &            &            &            &            & \checkmark \\
  One Sparse Operand              &            & \checkmark &            & \checkmark &            &\checkmark \\
  Multiple Sparse  Operands                 &            & \checkmark &            &            &            &\checkmark \\
  Run-length               &            &            &            &            &            & \checkmark \\
  Symmetric                &            &            &            &            & \checkmark & \checkmark \\
  Regular Sparse Blocks    &            & \checkmark &            &            &            & \checkmark \\
  Irregular Sparse Blocks  &            &            &            &            &            &\checkmark \\
  Ragged                   &            &            & \checkmark &            &            & \checkmark \\
  \end{tabular}
  \caption{Support for various data structures across tools. Finch supports \textbf{both} complex programs and complex data structures.}
  \label{tab:data_structures}
\end{minipage}
\end{table}

>>>>

%Array programming is the core abstraction behind many of the modern miracles of computing (e.g. neural networks, scientific simulation, database processing). 
Arrays are the most fundamental abstraction in computer science. Arrays and lists are often the first-taught datastructure
\cite[Chapter 2.2]{abelson_structure_1996}, \cite[Chapter 2.2]{knuth_art_1997}.
%
Arrays are also universal across programming languages, from their introduction
in Fortran in 1957 to present-day languages like Python
\cite{backus_fortran_1957}, keeping more-or-less the same semantics.
%
Modern
array programming languages such as NumPy, SciPy, MatLab, TensorFlow, PyTorch,
and Halide have pushed the limits of productive data processing with arrays,
fueling breakthroughs in machine learning, scientific computing, image
processing, and more  \cite{harris_array_2020, virtanen_scipy_2020,
moler_history_2020, abadi_tensorflow_2016,
paszke_pytorch_2019,ragan-kelley_halide_2013}\saman{Put these citations next to each language - otherwise associated with the breakthroughs...}.
% EASY CUT
%These frameworks have been the
%subject of extensive industry investment to enable performant implementations,
%and often operate at the peak capacity of the hardware they run on
%\cite{lo_roofline_2015}.

The success and ubiquity of arrays is largely due to their simplicity. 
%
Since their introduction, multidimensional arrays have represented dense, rectilinear,
integer grids of points. 
%
By \textbf{dense}, we mean that indices are mapped to value via a simple formula relating multidimensional space to linear memory.
%
Consequently, dense arrays offer extensive compiler optimizations and many convenient interfaces.
%
%This
%simplicity enables extensive interoperability, convenience layers, and
%optimizations by breaking the abstraction barrier between array representation
%and array storage.  
%
Compilers understand dense array computations across many
programming constructs, such as for and while loops, breaks, parallelism,
caching, prefetching, multiple outputs, scatters, gathers, vectorization,
loop-carry-dependencies, and more. Several\saman{myriad of} optimizations have been developed for
dense arrays, such as loop fusion, loop tiling, loop unrolling, and loop
interchange.
%
However, while dense arrays are the easiest way to program for performance, the world is not all dense.

@@

%
\textbf{Currently, support for structured data is fragmented and incomplete}.
%
Experts must hand write variations of even the simplest kernels, like matrix
multiply, for each data structure/data set and architecture to get performance.
%
Implementations must choose a small set of features to support well, resulting
in a compromise between \textbf{program flexibility} and \textbf{data structure
flexibility}.
%
Hand-written solutions are collected in diverse libraries like
MKL, OpenCV, LAPACK or SciPy~\cite{ bradski2000opencv, anderson1999lapack, virtanen2020scipy, psarras2022linear}. 
%
However, libraries will only ever support a subset of
programs on a subset of data structure combinations.
%
Even the most advanced
libraries, such as the GraphBLAS, which support a wide variety of sparse
operations over various semi-rings always lack support for other features, such
as tensors, fused outputs, or runs of repeated values~\cite{bulucc2017design, mattson2019lagraph}.
%
While dense array
compilers support an enormous variety of program constructs like early break and
multiple left hand sides, they only support dense arrays~\cite{ragan-kelley_halide_2013,grosser2012polly}.  
%
Special-purpose
compilers like TACO, Taichi, StructTensor, or CoRa which support a select subset of structured data
structures (only sparse, or only ragged arrays) must compromise by greatly
constraining the classes of programs which they support, such as tensor
contractions \cite{kjolstad_tensor_2019, hu_taichi_2019, ghorbani2023compiling, fegade_cora_2022}. \saman{when you cite multiple projects, put the citations next to each project.}
%
This trade-off is visualized in Tables \ref{tab:features} and \ref{tab:data_structures}.
%

@@

\begin{table}[h!]
\noindent
\flushleft
\begin{minipage}[t]{.49\textwidth}
  \scriptsize
  \begin{tabular}{l|cccccc}
  \textbf{Feature / Tool} & \rothead{Halide} & \rothead{Taco} & \rothead{Cora} & \rothead{Taichi} & \rothead{Stur} & \rothead{Finch} \\
  \hline
  Einsums/Contractions & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
  % High-Level API           &            & \checkmark &            &            &            & \checkmark \\
  % Automatic API Fusion     &            &            &            &            &            & \checkmark \\
%  Parallelism             & \checkmark & \checkmark & \checkmark & \checkmark&           & \checkmark \\
  Multiple LHS             & \checkmark &            & \checkmark & \checkmark &            & \checkmark \\
  Affine Indices           & \checkmark &            &            & \checkmark & \checkmark & \checkmark \\
  Recurrence               & \checkmark &            &            &            &            &           \\
  If-Conditions and Masks  & \checkmark & \checkmark &            & \checkmark &            & \checkmark \\
  Scatter Gather           & \checkmark &            &            & \checkmark &            &\checkmark \\
  Early Break              &            & \checkmark &            & \checkmark &            &\checkmark \\
  Unrestricted Read/Write              &    \checkmark        &  &            &  &            &  \\
  \end{tabular}
  \caption{Feature support across various tools.}
  \label{tab:features}
\end{minipage} 
\begin{minipage}[t]{.49\textwidth}
  \flushright
  \scriptsize
  \begin{tabular}{l|cccccc}
  \textbf{Feature / Tool} & \rothead{Halide} & \rothead{Taco} & \rothead{Cora} & \rothead{Taichi} & \rothead{Stur} & \rothead{Finch} \\
  \hline
  Dense                    & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
  Padded                   & \checkmark &            &            &            &            & \checkmark \\
  One Sparse Operand              &            & \checkmark &            & \checkmark &            &\checkmark \\
  Multiple Sparse  Operands                 &            & \checkmark &            &            &            &\checkmark \\
  Run-length               &            &            &            &            &            & \checkmark \\
  Symmetric                &            &            &            &            & \checkmark & \checkmark \\
  Regular Sparse Blocks    &            & \checkmark &            &            &            & \checkmark \\
  Irregular Sparse Blocks  &            &            &            &            &            &\checkmark \\
  Ragged                   &            &            & \checkmark &            &            & \checkmark \\
  \end{tabular}
  \caption{Support for various data structures across tools. Finch supports \textbf{both} complex programs and complex data structures.}
  \label{tab:data_structures}
\end{minipage}
\vspace{-24pt}
\end{table}
